Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-13 MScDiplom_Lemenkova.tex
Comments of Valentijn: 8, 9, 11, 36 (RTM, upscale mapping, discussion of spectra, optical measurement techniques)

Understanding radiance distribution within a water colum\index[concepts]{Seawater optical properties!RTM}n is necessary for the studies of the underwater visibility, because scattering properties of the water body naturally vary with changing depth, wavelength and environmental conditions.  Simulation of the radiance quantities for natural water bodies enables to analyse seafloor color remote sensing properties. Therefore, the artificial modelling of the seawater optical properties by means of the \ac{RTM} is used when water optics is studied under changing environment (e.g., depths, sun angle, suspended particles in water column). \\ In such cases a retrieval of water optical parameters from the remote measurements should be tested and analyzed. 
We briefly discuss below the most effective RTM software and algorithms from a range of various best-known up-to-date tools. 
		\item [*] The \ac{6SV1} \index[concepts]{Seawater optical properties!RTM!6SV1} is a US atmospheric correction algorithm, developed in the University of Maryland; adjusted for the \ac{MODIS} satellite imagery lookup tables. The 6S algorithm has been \href{http://grass.osgeo.org/grass64/manuals/html64_user/i.atcorr.html}{implemented} by the \ac{GRASS} software for the atmospheric correction 				 
		\item [*] Another example of the RTM adjusted for a specific imagery is a German \ac{KOPRA}\index[concepts]{Seawater optical properties!RTM!KOPRA}, fitted for \ac{MIPAS} and \ac{ENVISAT} imagery.
		\item [*]  The \href{http://www.sequoiasci.com/products/Hydrolight.aspx}{\textit{HydroLight RTM}} commercial software is an advanced model for oceanic radiative transfer calculations, developed by the Sequoia Scientific, Inc., USA. It is designed to solve a wide range of problems in optical oceanography and limnology.   
		\item [*] The British \ac{RTTOV-9}\index[concepts]{Seawater optical properties!RTM!RTTOV-9} is a radiative transfer model, developed for nadir viewing atmospheric sounders and imagers. It includes a number of useful tolls, e.g. can compute sea-surface emissivity for each channel, enables, to specify cloudiness for radiance calculations, etc.  This program runs under Unix/Linux platforms and is open source. 
		\item [*] The \ac{RRTM}\index[concepts]{Seawater optical properties!RTM!RRTM} is one more RTM freeware, designed in the USA, Massachusetts. 
		\item [*] The \href{http://www.modtran.org/}{\textit{MODTRAN}}\index[concepts]{Seawater optical properties!RTM!MODTRAN}, a commercial software from Spectral Sciences Inc., is another example of the RTM. Among its newest updates is generation of atmospheric correction data; it also has an option to write spectral output in binary, and convert it to \ac{ASCII}.
		\item [*] A \ac{DISORT}\index[concepts]{Seawater optical properties!RTM!DISORT}  is another RTM freeware , developed as an algorithm for calculation radiance of the scattering and absorbing Earth's atmosphere and transmissivity computations within the waveband range from UV to radar.  
		\item [*] The \ac{CRTM} \index[concepts]{Seawater optical properties!RTM!CRTM} is designed by the  (\href{http://www.jcsda.noaa.gov/}{JCSDA}, \ac{NASA}) and includes Surface Emissivity/Reflectivity Models, Cloud Absorption/Scattering Model and Gaseous Absorption Model.
		\item [*] An \ac{4AOP}\index[concepts]{Seawater optical properties!RTM!4AOP} is a French RTM freeware, designed in cooperation by \href{http://www.noveltis.fr/fr/index.php}{Noveltis} and \href{http://www.cnes.fr/}{the French Space Agency}, \ac{CNES}. Being a fine program which enables accurate computations of the radiance, defining other emissivity functions and other set up, it is however, particularly designed for the infrared waves (thus, not quite suitable in our case). To the advantages of this freeware we would count its user friendly interface and open source accessibility for the academic purposes. 
For our purpose we have chosen the \ac{WASI}\index[concepts]{Seawater optical properties!RTM!WASI} \ac{RTM} software, due to its effectiveness, adaptability for the Mediterranean environment, open source availability, coverage of necessary wavebands and clear, user friendly interface, enabling to adjust various environmental parameters.

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}. \\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}). One of the examples of the acoustic sonar sensor designed for seabed classification is the British \href{http://sonavision.co.uk/products.asp?cat_id=1}{RoxAnn} system, which application for the seafloor mapping is \href{http://www.ozcoasts.org.au/geom_geol/toolkit/Tech_CA_asc.jsp}{very well described} on the \href{http://www.ozcoasts.org.au/}{OzCoasts website}.\\ However, the applications of the acoustic methods requires specialized expensive equipment and is mostly used in the deep open waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that, initially intended for the bathymetric surveying, acoustic equipment is mostly designed for the geomorphological and geological studies of the underwater substratum and are, therefore, more adjusted for the benthic habitat discrimination, and they are not effective for the identifying of the biological species composition or even the presence of aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}. The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.

@@

Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}. \\ Pages: $<$\pageref{page-17}$>$, $<$\pageref{page-21}$>$}
	\label{fig:13}
\end{wrapfigure}

However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. Integrated together with \ac{GIS} vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.
However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Mellors09a}\label{Mellors09a}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broadband multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}. Page: $<$\pageref{page-20}$>$}
	\label{fig:2.3}
\end{wrapfigure}

Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers (Fig.\ref{fig:2.1})\label{page-17}, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

\begin{table}[htbp]
	\caption{Characteristics of selected ocean-colour sensors. Page: $<$\pageref{page-47}$>$}
	\begin{center}
		\rowcolors{1}{OldLace}{FloralWhite}
		\begin{tabular}{|p{2cm}|p{20mm}|p{15mm}|p{17mm}|p{15mm}|p{15mm}|p{15mm}|}
			\hline\hline
			\textbf{Satellite Sensor:} & {\textbf{Landsat}} & {\textbf{SPOT}} & {\textbf{ENVISAT}} & {\textbf{Terra}} & {\textbf{GeoEye}} & {\textbf{Nimbus 7}} \\ \hline\hline
				\textbf{Origin} & USA & France & Europe (ESA) & USA (NASA) & USA & USA \\ \hline
				\textbf{Instrument} & \textit{TM, ETM+} & \textit{Pan, XS} & \textit{MERIS} & \textit{MODIS} & \textit{SeaWiFS} & \textit{CZCS}\\ \hline
			\textbf{Number of channels} & 8 & 6 & 15 & 36 & 8 & 5 \\ \hline
			\textbf{Wavelength coverage} & 185-1700 & 500-1750 & 412.5- 900 & 469-1640 & 412-865 & 443-750 \\ \hline
			\textbf{Ground resolution (nadir), m} & 15-pan,30(MS),60 (TIR) & 2.5, 5, 10,20 & 1.2 km/300 m & 1.0 km & 1.13 km & 825 m \\ \hline
			\textbf{Band Width (nm)} & 0.45-10.4 & 10,20 & 2.5,7.5,9,10 14,20 & 0.18,0.3,0.5,10 - 50 & 20,40 & 20,100 \\ \hline
			\textbf{Launched} & 1972, 1999 & 1986 & 2002 & 1999 & 1997 & 1978 \\ \hline
			\textbf{Recurrent period, days} & 16 & 24 & 35 & 16 & 16 & 16 \\ \hline
		\end{tabular}
	\end{center}
	\label{tab:1}
\end{table}

@@

Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historical mapping or environmental analysis of change detection of the seagrass
landscapes.
The \ac{Landsat TM}\index[concepts]{Remote sensing!multispectral imagery!Landsat TM} and \ac{Landsat ETM+}\index[concepts]{Remote sensing!multispectral imagery!Landsat ETM+} data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
\ac{Landsat TM} towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.
Another well-known multispectral sensor, \ac{SPOT}\index[concepts]{Remote sensing!multispectral imagery!SPOT} provides multispectral imagery with a spatial resolution of 10 m (Fig.\ref{fig:2.4}), covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \ac{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. 
The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}\index[concepts]{Remote sensing!multispectral imagery!IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility.

@@

Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index[concepts]{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates high suitability of \ac{CASI} images for the fine-scale mapping of the seagrass
landscapes. Thus, \ac{CASI} and Quickbird-2\index[concepts]{Remote sensing!multispectral imagery!Quickbird} images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.
\ac{ASTER}\index[concepts]{Remote sensing!multispectral imagery!ASTER}), launched in
1999 on board Terra sensor\index[concepts]{Remote sensing!multispectral imagery!Terra sensor}, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:2.4}). It has diverse
subsystems for the \ac{VNIR} with 15 m resolution, shortwave infrared \ac{SWIR}, and
\ac{TIR} wavelength regions \cite{Kalinowski04}\label{Kalinowski04}. For each channel there is separate
on board calibration \ac{OBC} system, telescope with independent pointing and different detector technology \cite{Arai05}\label{Arai05}. 
This makes \ac{ASTER}\label{page-19} imagery especially suitable for the application towards detailed
mapping of surface temperature, emissivity and reflectance of objects and bathymetric elevations as
well (Fig.\ref{fig:2.4}). Successful application of the \ac{ASTER} imagery towards mapping of submerged vegetation
reported, for example, by Hirose \cite{Hirose04}\label{Hirose04}. 
Other multispectral images have also been used for the
seagrass interpretation. For example, the success of using imagery from the multispectral airborne
scanner Daedalus AADS1268 is reported by Heege et al \cite{Heege03}\label{Heege03} where they aim at classification of
macrophytes in shallow waters of the Lake Constance.

>>>>

Understanding radiance distribution within a water colum\index[concepts]{Seawater optical properties!RTM}n is necessary for the studies of the underwater visibility, because scattering properties of the water body naturally vary with changing depth, wavelength and environmental conditions.  Simulation of the radiance quantities for natural water bodies enables to analyse seafloor color remote sensing properties. Therefore, the artificial modelling of the seawater optical properties by means of the \ac{RTM} is used when water optics is studied under changing environment (e.g., depths, sun angle, suspended particles in water column). \\ In such cases a retrieval of water optical parameters from the remote measurements should be tested and analyzed. 
\newline
We briefly discuss below the most effective RTM software and algorithms, suitable for underwater radiance simulations, from a range of various best-known up-to-date tools. 
	\begin{itemize}
		\item [*] The \ac{6SV1} \index[concepts]{Seawater optical properties!RTM!6SV1} is a US atmospheric correction algorithm, developed in the University of Maryland; adjusted for the \ac{MODIS} satellite imagery lookup tables. The 6S algorithm has been \href{http://grass.osgeo.org/grass64/manuals/html64_user/i.atcorr.html}{implemented} by the \ac{GRASS} software for the atmospheric correction 	 
		\item [*] Another example of the RTM adjusted for a specific imagery is a German \ac{KOPRA}\index[concepts]{Seawater optical properties!RTM!KOPRA}, fitted for \ac{MIPAS} and \ac{ENVISAT} imagery.
		\item [*]  The \href{http://www.sequoiasci.com/products/Hydrolight.aspx}{\textit{HydroLight RTM}} commercial software is an advanced model for oceanic radiative transfer calculations, developed by the Sequoia Scientific, Inc., USA. It is designed to solve a wide range of problems in optical oceanography and limnology.   
		\item [*] The British \ac{RTTOV-9}\index[concepts]{Seawater optical properties!RTM!RTTOV-9} is a radiative transfer model, developed for nadir viewing atmospheric sounders and imagers. It includes a number of useful tolls, e.g. can compute sea-surface emissivity for each channel, enables, to specify cloudiness for radiance calculations, etc.  This program runs under Unix/Linux platforms and is open source. 
		\item [*] The \ac{RRTM}\index[concepts]{Seawater optical properties!RTM!RRTM} is one more RTM freeware, designed in the USA, Massachusetts. 
		\item [*] The \href{http://www.modtran.org/}{\textit{MODTRAN}}\index[concepts]{Seawater optical properties!RTM!MODTRAN}, a commercial software from Spectral Sciences Inc., is another example of the RTM. Among its newest updates is generation of atmospheric correction data; it also has an option to write spectral output in binary, and convert it to \ac{ASCII}.
		\item [*] The \ac{CRTM} \index[concepts]{Seawater optical properties!RTM!CRTM} is designed by the  (\href{http://www.jcsda.noaa.gov/}{JCSDA}, \ac{NASA}) and includes Surface Emissivity/Reflectivity Models, Cloud Absorption/Scattering Model and Gaseous Absorption Model.
	\end{itemize}
For our purpose we have chosen the \ac{WASI}\index[concepts]{Seawater optical properties!RTM!WASI} \ac{RTM} software, due to its effectiveness, adaptability for the Mediterranean environment, open source availability, coverage of necessary wavebands and clear, user friendly interface, enabling to adjust various environmental parameters.

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}.  The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.\\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}). One of the examples of the acoustic sonar sensor designed for seabed classification is the British \href{http://sonavision.co.uk/products.asp?cat_id=1}{RoxAnn} system, which application for the seafloor mapping is \href{http://www.ozcoasts.org.au/geom_geol/toolkit/Tech_CA_asc.jsp}{very well described} on the \href{http://www.ozcoasts.org.au/}{OzCoasts website}.\\ However, the applications of the acoustic methods requires specialized expensive equipment and is mostly used in the deep open waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that, initially intended for the bathymetric surveying, acoustic equipment is mostly designed for the geomorphological and geological studies of the underwater substratum and are, therefore, more adjusted for the benthic habitat discrimination, and they are not effective for the identifying of the biological species composition or even the presence of aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}.\\
The current study is based on the application of the remote sensing optical measurement techniques, due to their effectiveness, non-destructive nature and availability of necessary tools: spectral radiometers and \ac{RTM}.

@@

Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.

\begin{wrapfigure}{r}{0.4\textwidth}[H]
	\centering
	\includegraphics[scale=0.35]{Fig-11.jpg}
	\caption{\textit{In situ} optical reflectance spectra of seagrass. Shaded areas - \% of spectra lying within the range of reflectance. 
		White lines - mean spectra. Source: \cite{Hochberg03b}\label{Hochberg03b}. \\ Pages: $<$\pageref{page-17}$>$, $<$\pageref{page-21}$>$}
	\label{fig:13}
\end{wrapfigure}

\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}. 
However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. 
Integrated together with \ac{GIS} vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.

\begin{wrapfigure}{r}{0.4\textwidth}[H]
	\centering
	\includegraphics[scale=0.25]{Fig-12.jpg}
	\caption{Difference between broadband multispectral and hyperspectral resolution of
		spectral signatures. Source: \cite{Clark99}\label{Clark99}. Page: $<$\pageref{page-20}$>$}
	\label{fig:2.3}
\end{wrapfigure}

However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Mellors09a}\label{Mellors09a}.
Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers (Fig.\ref{fig:2.1})\label{page-17}, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

\begin{table}[htbp]\footnotesize
	\caption{Characteristics of selected ocean-colour sensors. Page: $<$\pageref{page-47}$>$}
	\begin{center}
		\rowcolors{1}{OldLace}{FloralWhite}
		\begin{tabular}{|p{2cm}|p{20mm}|p{15mm}|p{17mm}|p{15mm}|p{15mm}|p{15mm}|}
			\hline\hline
			\textbf{Satellite Sensor:} & {\textbf{Landsat}} & {\textbf{SPOT}} & {\textbf{ENVISAT}} & {\textbf{Terra}} & {\textbf{GeoEye}} & {\textbf{Nimbus 7}} \\ \hline\hline
				\textbf{Origin} & USA & France & Europe (ESA) & USA (NASA) & USA & USA \\ \hline
				\textbf{Instrument} & \textit{TM, ETM+} & \textit{Pan, XS} & \textit{MERIS} & \textit{MODIS} & \textit{SeaWiFS} & \textit{CZCS}\\ \hline
			\textbf{Number of channels} & 8 & 6 & 15 & 36 & 8 & 5 \\ \hline
			\textbf{Wavelength coverage} & 185-1700 & 500-1750 & 412.5- 900 & 469-1640 & 412-865 & 443-750 \\ \hline
			\textbf{Ground resolution (nadir), m} & 15-pan,30(MS),60 (TIR) & 2.5, 5, 10,20 & 1.2 km/300 m & 1.0 km & 1.13 km & 825 m \\ \hline
			\textbf{Band Width (nm)} & 0.45-10.4 & 10,20 & 2.5,7.5,9,10 14,20 & 0.18,0.3,0.5,10 - 50 & 20,40 & 20,100 \\ \hline
			\textbf{Launched} & 1972, 1999 & 1986 & 2002 & 1999 & 1997 & 1978 \\ \hline
			\textbf{Recurrent period, days} & 16 & 24 & 35 & 16 & 16 & 16 \\ \hline
		\end{tabular}
	\end{center}
	\label{tab:1}
\end{table}

@@

Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historical mapping or environmental analysis of change detection of the seagrass
landscapes.
The \ac{Landsat TM}\index[concepts]{Remote sensing!multispectral imagery!Landsat TM} and \ac{Landsat ETM+}\index[concepts]{Remote sensing!multispectral imagery!Landsat ETM+} data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
\ac{Landsat TM} towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.

@@

Another well-known multispectral sensor, \ac{SPOT}\index[concepts]{Remote sensing!multispectral imagery!SPOT} provides multispectral imagery with a spatial resolution of 10 m (Fig.\ref{fig:2.4}), covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \ac{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. 
The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}\index[concepts]{Remote sensing!multispectral imagery!IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. 
Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index[concepts]{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.
